---
title: "Data Tidying with tidyr"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_depth: 1
    fontsize: 13pt
---
##

This is a reading note from Data Science with R. For more details this chapter please see <http://garrettgman.github.io/tidying/>.

## Prerequisites

DSR is the dataset.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_package}
#install.packages(c("tidyr","devtools"))
#devtools::install_github("garrettgman/DSR")
```

## Four types of data table

```{r}
library(DSR)
table1
table2
table3
table4
table5
```

# So, which dataset we will consider as a tidy dataset?

## Three rules of tidy datasets

Your data will be easier to work with R if it follows three rules:

1. Each **variable** in the dataset is placed in its own **column**
2. Each **observation** is placed in its own **row**
3. Each **value** is placed in its own **cell**

Data that satisfies these rules is known as tidy data. Notice that *table1* is tidy data.

# Why we need tidy data?
## Tidy data with R

Tidy data works well with R because R is a **vectorized** programming language.

1. Data structures in R are built from **vector**
2. R's operations are optimized to work with vectors.

Tidy data take advantage of both of these traits.

In R, a data frame is a list of vector that R displays as a table. When your data is tidy, the values of each vairable fall in their own column vector.

---

With tidy data, you can easily extract and manipulate the values of a variable:

```{r extract_var}
table1$cases
```

Then you can extract abd manipulate the value of variables in tidy data with concise, simple code:

```{r}
mean(table$cases)
table1$cases / table$population * 10000
```

